Follow-up Samples

### 2017-04-18 submission/manuscript.Rmd

bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk factor.

@@

This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [ADD REFERENCES FOR CAUSATION;  @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients we observed changes in their microbiota. The microbiota of patients diagnosed with carcinoma changed significant more than the other groups. This change resulted in communities that more closely resembled those of patients with normal colons.

* Does treatment affect the colonic microbiota in a predictable manner? For those with carcinoma

* Does the treatment alter the community to more closely resemble that of individuals with normal colons? For those with carcinoma.

* Heterogeneity of the population (common problem for all microbiome studies) made it impossible to identify speicfi microbiota that changed with treatment; however, RF made it possible to identify consortia of microbiota that were associated with treatment and recovery.

* End.

Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3]. Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.

Here we show that the microbiome changes towards normal after treatment for carcinoma

Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g. *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not. The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models. This suggests that members within the commensal community may be the first that change during CRC pathogenesis. These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

Many of the common OTUs that we identified taxonomically classified to potential butyrate producers (e.g. *Clostridiales*, *Roseburia*. and *Anaerostipes*) [Table S2]. Other OTUs classified to bacteria that are inhibited by polyphenols (e.g. *Bacteroides*). Both butyrate and polyphenols are thought to be protective against cancer, in part by reducing inflammation [@louis_gut_2014]. These protective compounds are derived from the breakdown of fiber, fruits, and vegetables by resident gut microbes. One example of this potential diet-microbiome-inflammation-polyp axis is that *Bacteroides*, which was highly prevalent in our models, are known to be increased in those with high non-meat based protein consumption [@zhu_intake_2016]. High protein consumption in general has been linked with an increased CRC risk [@mu_colonic_2016]. Conversely, *Bacteroides* are inhibited by polyphenols which are derived from fruits and vegetables [@ozdal_reciprocal_2016]. Our data fits with the hypothesis that the microbial metabolites from breakdown products within our own diet could not only help to shape the existing community but also have an effect on CRC risk and disease progression. Within this context the commensal community may be an important modifiable risk factor for monitoring and preventing CRC recurrence after treatment.

Building off of our results, an area for future research, is that within our study we do not know whether individuals who were still classified as positive by the carcinoma model eventually had a subsequent CRC recurrence. This information would help to strengthen the case for this model keeping numerous individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having carcinoma. It would also provide additional evidence that the microbiome could be used as a risk stratification tool in monitoring recurrence risk and whether different interventions could potentially change this community and lower the probability of future recurrence.

Despite the stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable difference in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment. Further, the ability for machine learning algorithms to take OTU data and successfully lower positive probability of carcinoma after treatment provides evidence that there are specific signatures, attributable to both inflammatory and resident commensal organisms, associated with treatment. Our data provides evidence that commensal bacteria may be important in the development of polyps, potential transition of advanced adenoma to carcinoma, and recovery of the microbiome in CRC following treatment.

>>>>

microbiome; colorectal cancer; polyps; treatment; risk factor.

@@

This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that the microbiome contains biomarkers to diagnosis colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Our study focused on comparing the microbiome of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients we observed changes in their microbiota. The microbiota of patients diagnosed with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with normal colons.

For the carcinoma group there was a larger change between their initial and follow up samples then for either adenoma or advanced adenoma.  This change was directionaly the same for each individual with carcinoma.  This observed change has  important implications for future research into CRC recurrence.  Knowing that the community changes towards normal after treatment means that it is reasonable to hypothesize that the microbiome could be a good biomarker for stratifying individuals between those most and least at risk for CRC recurrence.  Interestingly, our `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability fits nicely within the range quoted for CRC recurrence (20-30%) [@hellinger_reoperation_2006; @ryuk_predictive_2014].

A potential limitation, within our study, is that we do not know whether individuals who had an increased positive probability for carcinoma on follow up eventually had a subsequent CRC recurrence.  This information would help to strengthen the case for the ability of our model to be used as a disease stratification tool for CRC recurrence. This would be an exciting area of future research that would be able to link treatment changes within the microbiome and CRC recurrence together.

The microbiome changes associated with carcinoma were found to be towards those with normal colons.  This would suggest that treatment for CRC is not only successful for removing the carcinoma but also at removing bacteria associated with them.  This is important due to the numerous studies that have pointed towards a link between the community and tumorgenesis [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].  The importance of the community to this process is further highlighted in our data [Figure 2, 4, & S3].  Interestingly, those who had an increase in their positive probability of carcinoma had no change in their median value for oral pathogens found within the carcinoma model but had a decrease in non-oral pathogens [Figure S4].  Although this finding was not statistically significant, this trend would further highlight the importance of community to tumorgenesis.  Future studies with larger power may be able to confirm these reported observations.

Although we were not able to identify specific OTUs due to the heterogeneity of the population.  This is a common problem that occurs in other fields of microbiome study [@sze_looking_2016].  This made it very difficult to identify individual OTUs associated with treatment and recovery towards a normal colon.  To overcome this large individual heterogeneity we used RF models that made it easier to identify consortia of microbiota that were associated with treatment and recovery.  A potential future direction would be to study whether specific gene signatures associated with these consortia of bacteria have a role to play in the recurrence of CRC.

Despite some of these stated shortcomings our findings add to the existing scientific knowledge on CRC and the microbiome: That there is a measurable change in the bacterial community after adenoma, advanced adenoma, or carcinoma treatment. Further, individuals with carcinoma had a larger change than adenoma or advanced adenoma indivdiuals and this change in community was towards those with normal colons. Our data provides additional evidence on the importance of the community  on recovery of the microbiome after CRC treatment and opens interesting avenues of research into how these changes may affect recurrence.

### 2017-04-18 submission/manuscript.tex

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subject in each group,
we restricted our models to only incorporate 10 OTUs. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
the 10 OTUs that were used for each model had little overlap with each
other {[}Figure 2{]}. These results support the earlier community-wide
analysis where we observed that the treatment had an impact on the
overall community structure; however, the effect of treatment was not
consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma
share}} Next, we determined whether treatment changed the microbiota in
a way that the post-treatment communities resembled that of patients
with normal colons. To test this, we used an expanded cohort of 423
individuals that were diagnosed under the same protocol as having normal
colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
We then constructed Random Forest models to classify the pre and
post-treatment samples as having their original diagnosis or having a
normal colon. The models performed well (adenoma AUC=0.62 - 0.72,
advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
The OTUs that were most important in the models for differentiating
between patients with normal colons and those with adenomas or advanced
adenomas were dominated by populations commonly considered as commensals
(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figure S3{]}. Although many of these OTUs were also
included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}Figure S3{]}. Finally, we applied these three
models to the pre and post-treatment samples for each diagnosis group
and quantified the change in the positive probability of the model. A
decrease in the positive probability would indicate that the colon more
closely resembled that of a patient with a normal colon. There was not a
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 3{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
that were initially diagnosed with a carcinoma had a higher positive
probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differs and how it might be used as an early screening tool
{[}23--27{]}. Here we show that the microbiome changes towards normal
after treatment for carcinoma and that chemotherapy and radiation did
not provide any additional change. Although some of the important OTUs
classified to genera from bacteria considered the ``usual suspects''
(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
many did not. The majority of important OTUs had taxonomic
classifications for resident gut microbes and were common for the
adenoma, advanced adenoma, and carcinoma models. This suggests that
members within the commensal community may be the first that change
during CRC pathogenesis. These subtle changes, in turn, could be the
first step in allowing more inflammatory bacteria to gain a foothold
within the colon {[}13{]}.

Unlike previous studies on the microbiome and CRC, ours focuses on
comparing the microbiome during recovery from treatment in adenoma,
advanced adenoma, and carcinoma groups. Although there were differences
for genera associated with specific bacterium linked with CRC {[}Figure
S3{]}, most of the important OTUs taxonomically classified to commensal
bacteria {[}Figure S3{]}. Although these changes may be subtle, due to
the lack of significant differences in the bacterial community pre- and
post-treatment in adenoma and advanced adenoma {[}Figure 1{]}; they
support the hypothesis that the first members of the community to change
and potentially stay changed even after treatment are those that are
commensal bacteria.

Many of the common OTUs that we identified taxonomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}. and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}28{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}29{]}. High protein consumption in general has been linked with an
increased CRC risk {[}30{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}31{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression. Within this context the commensal community may
be an important modifiable risk factor for monitoring and preventing CRC
recurrence after treatment.

A limitation in our study was that there was a significant difference in
the time elapsed in the collection of the post-treatment sample between
adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
with time passed being less for adenoma (255 ± 42 days) and advanced
adenoma (250 ± 41) than carcinoma (351 ± 102). These results would
indicate that some of the differences observed between the carcinoma and
adenoma groups could be due to differences in collection time.
Specifically, it could confound the observation that carcinomas changed
more than adenomas {[}Figure 1{]}. However, there are two reasons that
this may not be the case. First, the advanced adenoma group had a higher
dissimilarity than adenoma but lower dissimilarity then carcinoma and
the collection time to their follow up samples was less than the adenoma
group. Second, this confounding would not affect the observations where
models were used since they were built using a different cohort
{[}Figures 2 \& S2-S3{]}.

Another limitation of this study was that it drew heavily from those
with Caucasian ancestry making it possible that the observations may not
be representative of those with either Asian or African ancestry.
Although our training and test set are relatively large we still run the
risk of over-fitting or having a model that may not be representative of
other populations. We have done our best to safeguard against this by
not only running 10-fold cross validation but also having over 100
different 80/20 splits to try and mimic the type of variation that might
be expected to occur.

Building off of our results, an area for future research, is that within
our study we do not know whether individuals who were still classified
as positive by the carcinoma model eventually had a subsequent CRC
recurrence. This information would help to strengthen the case for this
model keeping numerous individuals above the cutoff threshold even
though at follow up they were diagnosed as no longer having carcinoma.
It would also provide additional evidence that the microbiome could be
used as a risk stratification tool in monitoring recurrence risk and
whether different interventions could potentially change this community
and lower the probability of future recurrence.

Despite the stated shortcomings our findings add to the existing
scientific knowledge on CRC and the microbiome: That there is a
measurable difference in the bacterial community after adenoma, advanced
adenoma, or carcinoma treatment. Further, the ability for machine
learning algorithms to take OTU data and successfully lower positive
probability of carcinoma after treatment provides evidence that there
are specific signatures, attributable to both inflammatory and resident
commensal organisms, associated with treatment. Our data provides
evidence that commensal bacteria may be important in the development of
polyps, potential transition of advanced adenoma to carcinoma, and
recovery of the microbiome in CRC following treatment.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}23{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonoscopy community within this time period
{[}32{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}33{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}33{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Model Building:}} The Random Forest {[}34{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals. There were a total of 239 individuals in the adenoma model,
262 individuals in the advanced adenoma model, and 266 indivdiuals in
the carcinoma model {[}Table 1{]}. Each model was then applied to our
67-person cohort testing prediction of adenoma intial (adenoma n = 22)
versus adenoma follow up (adenoma n = 0), advanced adenoma initial
(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
follow up (carcinoma n = 1).

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}35{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtining the median
MDA from the 100 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\hypertarget{ref-louis_gut_2014}{}
28. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
metabolites and colorectal cancer. Nature Reviews Microbiology
{[}Internet{]}. 2014 {[}cited 2017 Feb 14{]};12:661--72. Available from:
\url{http://www.nature.com/doifinder/10.1038/nrmicro3344}

\hypertarget{ref-zhu_intake_2016}{}
29. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
Proteins Substantially Increased the Relative Abundance of Genus
Lactobacillus in Rat Feces. PloS One. 2016;11:e0152678.

\hypertarget{ref-mu_colonic_2016}{}
30. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
Compared with a Normal-Protein Diet. The Journal of Nutrition.
2016;146:474--83.

\hypertarget{ref-ozdal_reciprocal_2016}{}
31. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
Reciprocal Interactions between Polyphenols and Gut Microbiota and
Effects on Bioaccessibility. Nutrients {[}Internet{]}. 2016 {[}cited
2017 Feb 14{]};8:78. Available from:
\url{http://www.mdpi.com/2072-6643/8/2/78}

\hypertarget{ref-obrien_impact_2013}{}
32. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-kozich_development_2013}{}
33. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

\hypertarget{ref-breiman_random_2001}{}
34. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
35. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-benjamini_controlling_1995}{}
36. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.

>>>>

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subject in each group,
we restricted our models to only incorporate 10 OTUs to limit the
likelihood that that models would overfit the data. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
the 10 OTUs that were used for each model had little overlap with each
other {[}Figure 2{]}. These results support the earlier community-wide
analysis where we observed that the treatment had an impact on the
overall community structure; however, the effect of treatment was not
consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma
share}} Next, we determined whether treatment changed the microbiota in
a way that the post-treatment communities resembled that of patients
with normal colons. To test this, we used an expanded cohort of 423
individuals that were diagnosed under the same protocol as having normal
colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
We then constructed Random Forest models to classify the pre and
post-treatment samples as having their original diagnosis or having a
normal colon. The models performed well (adenoma AUC=0.62 - 0.72,
advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
The OTUs that were incorporated into the adenoma and advanced adenoma
models largely overlapped and those OTUs that were used to classify the
carcinoma samples were largely distinct from those of the other two
models {[}Figure 4A{]}. Among the OTUs that were shared across the three
models were those populations commonly considered as commensals (e.g.
\emph{Faecalibacterium}, \emph{Lachnospiraceae}, \emph{Bacteroides},
\emph{Dorea}, \emph{Anaerostipes}, and \emph{Roseburia}) {[}Figures
4B{]}. Although many of these OTUs were also included in the model
differentiating between patients with normal colons and those with
carcinoma, this model also included OTUs affiliated with populations
that have previously been associated with carcinoma
(\emph{Fusobacterium}, \emph{Porphyromonas}, \emph{Parvimonas})
{[}Figure S3{]}. Finally, we applied these three models to the pre and
post-treatment samples for each diagnosis group and quantified the
change in the positive probability of the model. A decrease in the
positive probability would indicate that the colon more closely
resembled that of a patient with a normal colon. There was not a
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 3{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
that were initially diagnosed with a carcinoma had a higher positive
probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon

@@

This study builds upon previous work that demonstrated a role for the
microbiota in tumorigenesis and that the microbiome contains biomarkers
to diagnosis colonic lesions {[}5--9,23--27{]}. Our study focused on
comparing the microbiome of patients diagnosed with adenoma, advanced
adenoma, and carcinoma before and after treatment. For all three groups
of patients we observed changes in their microbiota. The microbiota of
patients diagnosed with carcinoma changed significantly more than the
other groups. This change resulted in communities that more closely
resembled those of patients with normal colons.

For the carcinoma group there was a larger change between their initial
and follow up samples then for either adenoma or advanced adenoma. This
change was directionaly the same for each individual with carcinoma.
This observed change has important implications for future research into
CRC recurrence. Knowing that the community changes towards normal after
treatment means that it is reasonable to hypothesize that the microbiome
could be a good biomarker for stratifying individuals between those most
and least at risk for CRC recurrence. Interestingly, our 23.08\%
prevalence of increased carcinoma probability fits nicely within the
range quoted for CRC recurrence (20-30\%) {[}3,4{]}.

A potential limitation, within our study, is that we do not know whether
individuals who had an increased positive probability for carcinoma on
follow up eventually had a subsequent CRC recurrence. This information
would help to strengthen the case for the ability of our model to be
used as a disease stratification tool for CRC recurrence. This would be
an exciting area of future research that would be able to link treatment
changes within the microbiome and CRC recurrence together.

The microbiome changes associated with carcinoma were found to be
towards those with normal colons. This would suggest that treatment for
CRC is not only successful for removing the carcinoma but also at
removing bacteria associated with them. This is important due to the
numerous studies that have pointed towards a link between the community
and tumorgenesis {[}10--12{]}. The importance of the community to this
process is further highlighted in our data {[}Figure 2, 4, \& S3{]}.
Interestingly, those who had an increase in their positive probability
of carcinoma had no change in their median value for oral pathogens
found within the carcinoma model but had a decrease in non-oral
pathogens {[}Figure S4{]}. Although this finding was not statistically
significant, this trend would further highlight the importance of
community to tumorgenesis. Future studies with larger power may be able
to confirm these reported observations.

Although we were not able to identify specific OTUs due to the
heterogeneity of the population. This is a common problem that occurs in
other fields of microbiome study {[}28{]}. This made it very difficult
to identify individual OTUs associated with treatment and recovery
towards a normal colon. To overcome this large individual heterogeneity
we used RF models that made it easier to identify consortia of
microbiota that were associated with treatment and recovery. A potential
future direction would be to study whether specific gene signatures
associated with these consortia of bacteria have a role to play in the
recurrence of CRC.

Despite some of these stated shortcomings our findings add to the
existing scientific knowledge on CRC and the microbiome: That there is a
measurable change in the bacterial community after adenoma, advanced
adenoma, or carcinoma treatment. Further, individuals with carcinoma had
a larger change than adenoma or advanced adenoma indivdiuals and this
change in community was towards those with normal colons. Our data
provides additional evidence on the importance of the community on
recovery of the microbiome after CRC treatment and opens interesting
avenues of research into how these changes may affect recurrence.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}23{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonoscopy community within this time period
{[}29{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}30{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}30{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Model Building:}} The Random Forest {[}31{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals. There were a total of 239 individuals in the adenoma model,
262 individuals in the advanced adenoma model, and 266 indivdiuals in
the carcinoma model {[}Table 1{]}. Each model was then applied to our
67-person cohort testing prediction of adenoma intial (adenoma n = 22)
versus adenoma follow up (adenoma n = 0), advanced adenoma initial
(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
follow up (carcinoma n = 1).

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}32{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}33{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtining the median
MDA from the 100 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\hypertarget{ref-sze_looking_2016}{}
28. Sze MA, Schloss PD. Looking for a Signal in the Noise: Revisiting
Obesity and the Microbiome. mBio. 2016;7.

\hypertarget{ref-obrien_impact_2013}{}
29. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-kozich_development_2013}{}
30. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

\hypertarget{ref-breiman_random_2001}{}
31. Breiman L. Random Forests. Machine Learning {[}Internet{]}. 2001
{[}cited 2013 Feb 7{]};45:5--32. Available from:
\href{http://link.springer.com/article/10.1023/A\%3A1010933404324\%20http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}{http://link.springer.com/article/10.1023/A\%3A1010933404324 http://link.springer.com/article/10.1023\%2FA\%3A1010933404324?LI=true}

\hypertarget{ref-anderson_permanova_2013}{}
32. Anderson MJ, Walsh DCI. PERMANOVA, ANOSIM, and the Mantel test in
the face of heterogeneous dispersions: What null hypothesis are you
testing? Ecological Monographs {[}Internet{]}. 2013 {[}cited 2017 Jan
5{]};83:557--74. Available from:
\url{http://doi.wiley.com/10.1890/12-2010.1}

\hypertarget{ref-benjamini_controlling_1995}{}
33. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
practical and powerful approach to multiple testing. Journal of the
Royal Statistical Society. Series B (Methodological). 1995;57:289--300.
